---
title: "Hannah Bensen - Investigative Reporting Workshop"
output: html_document
date: "2025-05-09"
---

A demo notebook for Dow Jones News Fund data interns

During the week’s training, you will work on a project called Investigate Your Community to produce detailed story pitches relevant to their summer internship positions. The goal is to have the students research a topic relevant to your newsroom so they can get to work immediately at the internship. 

Here are some examples from last year’s DJNF data journalism training:

Hannah Rosenberger, at ASI Media/Counselor magazine, produced this analysis on name, image and likeness trends: https://hrosenberger.github.io/ASI-NIL-Project/

Shifra Dayak, an intern at the Connecticut Mirror, produced this analysis of environment spill data: https://shifradayak.github.io/DJNF_Merrill/ct_env_spills.html.

Michael Zhang, an intern at the Houston Landing, explored local election data: https://michaelzhangreporter.github.io/Houston_Campaign_Finance/


Here's what your newsroom suggested for a project
I'm afraid we don't have a specific data set or topic for Hannah Bensen to examine. We're continuing to work on projects related to long-term health effects of lead and government accountability related to cleanup regarding lead paint. We also have several police-use-of-force stories underway and are nearing completion on one related to police-surveillance tools and analytics. Aarushi has talked to Hannah about the pitch process and our interests so Hannah may be looking into a few ideas. 



Blood Lead Levels (µg/dL) among U.S. Children <72 Months of Age, by State, Year, and Blood Lead Level (BLL)

This notebook helps you get started

# Load software libraries
```{r}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("readxl")
#install.packages("rvest")
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)

```


# Load your data

2024

```{r}

# 
# YOUR_DATA <- read.csv("../data/XXXXX.csv") |> 
#     clean_names()

data <- read_excel("../data/cbls-national-data-508.xlsx", skip=2) |> 
     clean_names()



```

Data details:
https://www.cdc.gov/lead-prevention/php/data/national-surveillance-data.html
BLL - Blood Lead Level, SD - Indicates non-zero counts fewer than 6 have been suppressed to protect confidentiality													
† - Partial annual data submission													
* - Indicates program did not receive childhood lead poisoning prevention funding from CDC													
‡ - Indicates data not currently available													
ǁ - Includes all popluation data in table													
Population estimates data source: CDC NCHS Bridged Race Single Year of Age population (U.S. Census Populations With Bridged Race Categories https://www.cdc.gov/nchs/nvss/bridged_race.htm#data).													
 - For years 2017-2019, the data use each respective Vintage year for the denominator data (e.g., 2017 data uses the 2017 Vintage bridged race postcensal population estimates).													
 - For years 2020-2021, due to data availability at the time of processing, Vintage 2020 was used for data years 2020 and 2021. 													
 - This will be updated with future data updates to align with pre-2020 years (e.g., 2021 data will use the 2021 Vintage postcensal population estimates, etc.).													
The current Blood Lead Reference Value of 3.5 µg/dL was enacted by CDC in October 2021. Categorizations of blood lead levels above and below this value will be used in blood lead surveillance tables starting with 2022 data. 				


From ChatGPT:
According to the Centers for Disease Control and Prevention (CDC), any blood lead level (BLL) of 5 µg/dL or higher is considered elevated and requires public health actions.

Here's a summary of the danger zones based on the image and your description:

BLLs ≥5 µg/dL: This is the threshold where lead exposure becomes concerning and may require intervention.
BLLs ≥10 µg/dL: This level indicates more significant lead exposure and is considered dangerous.
BLLs 5-9 µg/dL: While lower than 10 µg/dL, this range is still concerning and above the CDC's reference level.
BLLs 10-14 µg/dL: This is a dangerous zone and indicates elevated lead exposure.
BLLs 15-19 µg/dL: This level is considered dangerous and requires intervention.
BLLs 20-24 µg/dL: This is a dangerous level of lead exposure.
BLLs 25-44 µg/dL: This is a significantly dangerous level and poses serious health risks.
BLLs ≥45 µg/dL: This is extremely dangerous and requires immediate medical attention.

# Explore the data types
```{r}

#glimpse(YOUR_DATA)

glimpse(data)

```

# Data exploration: states with most dangerous levels

```{r}
data |> 
  select(year, state, number_of_children_with_confirmed_bl_ls_45_mg_d_l) |> 
  filter(number_of_children_with_confirmed_bl_ls_45_mg_d_l >0 ) |> 
  arrange(desc(year))
```

# Questions 

Write down questions you want answered:



# Data Details
I'd consider converting the SD to 1 and running a sum for all levels above 10 ug/dl.
"SD - Indicates non-zero counts fewer than 6 have been suppressed to protect confidentiality." So SD equals something between 1 and 6. Converting to 1 would be a conservative estimate 		
